Pretty darn good control: when are approximate solutions better than approximate models

### 2023-01-31 manuscript/manuscript.Rmd
mini changes to math models section, notation update for beginning of RL section

Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (saddle-node bifurcation) (see Fig. \ref{fig:may}).
Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $\hat X_t$ of the system, these two attractors separated by an unstable fixed point.
At a certain threshold value of $H$, however, the top stable fixed point collides with the unstable fixed point and both are annihilated.
For this value of $H$, and for higher values, only the lower fixed point remains.
This also creates the phenomenon of *hysteresis*, where returning $H$ to its original value is not sufficient to restore $\hat X_t$ to the original stable state.

@@

Consider the following three species *generalization* of eq. \eqref{eq:may}:

\begin{align}
\label{eq:3d model}
\begin{split}
  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t,\\
  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
  \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t.
\end{split}
\end{align}
The model is conceptualized in Fig. \ref{fig:conceptual}, panel B.
It contains three populations, the state at time $t$ is $(X_t, Y_t, Z_t)$. 
Species $Z$ preys on both $X$ and $Y$, while the latter two compete for resources.
There are ten parameters in this model: 
The growth rate and carrying capacity, $r_X$, $K_X$, $r_Y$ and $K_Y$, of $X$ and $Y$.
A parameter $c_{XY}$ mediating a Lotka-Volterra competition between $X$ and $Y$.
A maximum predation rate $\beta$ and a predation rate half-maximum biomass $c$ specifying how $Z$ forages on $X$ and $Y$.
A parameter $D$ regulating a relative preference of $Z$ to prey on $Y$.
Finally, a death rate $d_Z$ and a parameter $f$ related to the birth rate of $Z$.

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/stX.png}
    \caption{$X$ component of fixed points.}
    \label{fig:stx}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/stY.png}
    \caption{$Y$ component of fixed points.}
    \label{fig:sty}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/stZ.png}
    \caption{$Z$ component of fixed points.}
    \label{fig:stz}
  \end{subfigure}
  \caption{Numerically-computed fixed points of the three-species deterministic model, eq. \eqref{eq:3d model}.
  Here, the parameter $D$ was varied between 0.75 and 1.25, and the three components of the fixed points are plotted separately, as a function of $D$.}
  \label{fig:fixed points}
\end{figure}

Throughout most of this paper we will fix the parameters of the model according to eq. \eqref{eq:param vals}.
At these values, the fixed points of the model are:\footnote{
We have included only three decimal digits for readability.
These fixed points may be generated by running the ‘‘stable\_points.py’’ script in \url{https://github.com/boettiger-lab/approx-model-or-approx-soln}.
}
\begin{align*}
  (X,Y,Z) \approx &(0.839, 0.054, 0.377), \quad 
    (0.048, 0.774, 0.480), \quad
    (0.359, 0.491, 0.509), \\
    &(0.666, 0.666, 0.000),\quad
    (0.899, 0.000, 0.303), \quad
    (0.000, 0.818, 0.457), \\
    &(1.000,0.000,0.000),\quad (0.000,1.000,0.000),\quad (0.000,0.000,0.000).
\end{align*}
The classification of fixed points for multi-dimensional systems is generally more complicated than in one dimension.
A fixed point might be stable against perturbations in one direction, but unstable against perturbations along another.
While a detailed classification of these fixed points is beyond the scope of this work, we point out that the point $(0.839, 0.054, 0.377)$ is stable against perturbations in all directions.
We will thus use it as a starting point for our simulations.

@@

We have, up to now, presented eqs. \eqref{eq:may} and \eqref{eq:3d model} as *deterministic* dynamical models for the sake of clarity.
In practice, we will consider a stochastic contribution to these dynamics.
This way, instead of eq. \eqref{eq:may}, we consider the following stochastic single-species system
\begin{align}
  \label{eq:may stoch}
  \Delta X_t
  = L(X_t;\ r, K) - F(X_t, H; \beta,c) - h_t + \eta_t,
\end{align}
where $\{\eta_t\}_t$ are independent Gaussian variables with zero mean.
The variance $\sigma^2$ of $\eta_t$ is an additional parameter of the model.

Similarly, we use the following dynamical model in place of eq. \eqref{eq:3d model}
\begin{align}
  \label{eq:3d model stoch}
  \begin{split}
    \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t + \eta_{X,t},\\
    \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t + \eta_{Y,t},\\
    \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t + \eta_{Z,t}.
\end{split}
\end{align}
Here, again, $\{\eta_{i,t}\}$ is a set of independent Gaussian variables centered at zero with variances, respectively, $\sigma_X^2$, $\sigma_Y^2$ and $\sigma_Z^2$.

@@

For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state being a vector of all the fish populations, $S=(V_1, V_2, H)$.
At each time step, the agent harvests one of the populations, $V_1$. 
This changes the state as
\begin{align*}
  (V_1,\ V_2,\ H) \mapsto ((1-q)V_1,\ V_2,\ H),
\end{align*}
where $q$ is a fishing *quota* set by the agent. 
This secures a reward of $qV_1$.
Afterwards, the environment undergoes a timestep under its natural dynamics given by \eqref{eq:3d model}.

>>>>

Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (saddle-node bifurcation) (see Fig. \ref{fig:may}).
Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $X_t$ of the system, these two attractors separated by an unstable fixed point.
At a certain threshold value of $H$, however, the top stable fixed point collides with the unstable fixed point and both are annihilated.
For this value of $H$, and for higher values, only the lower fixed point remains.
This also creates the phenomenon of *hysteresis*, where returning $H$ to its original value is not sufficient to restore $X_t$ to the original stable state.

@@

Consider the following three species generalization of eq. \eqref{eq:may}:

\begin{align}
\label{eq:3d model}
\begin{split}
  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t,\\
  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
  \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t.
\end{split}
\end{align}
The model is conceptualized in Fig. \ref{fig:conceptual}, panel B.
It contains three populations, the state at time $t$ is $(X_t, Y_t, Z_t)$. 
Species $Z$ preys on both $X$ and $Y$, while the latter two compete for resources.
There are ten parameters in this model: 
The growth rate and carrying capacity, $r_X$, $K_X$, $r_Y$ and $K_Y$, of $X$ and $Y$.
A parameter $c_{XY}$ mediating a Lotka-Volterra competition between $X$ and $Y$.
A maximum predation rate $\beta$ and a predation rate half-maximum biomass $c$ specifying how $Z$ forages on $X$ and $Y$.
A parameter $D$ regulating a relative preference of $Z$ to prey on $Y$.
Finally, a death rate $d_Z$ and a parameter $f$ scaling the birth rate of $Z$.

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/stX.png}
    \caption{$X$ component of fixed points.}
    \label{fig:stx}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/stY.png}
    \caption{$Y$ component of fixed points.}
    \label{fig:sty}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/stZ.png}
    \caption{$Z$ component of fixed points.}
    \label{fig:stz}
  \end{subfigure}
  \caption{Numerically-computed fixed points of the three-species deterministic model, eq. \eqref{eq:3d model}.
  Here, the parameter $D$ was varied between 0.75 and 1.25, and the values of the other parameters are fixed as in eq. \eqref{eq:param vals}. 
  The three components of the fixed points are plotted separately, as a function of $D$.}
  \label{fig:fixed points}
\end{figure}

Throughout most of this paper we will fix the parameters of the model according to eq. \eqref{eq:param vals}.
At these values, the fixed points of the model are:\footnote{
We have included only three decimal digits for readability.
These fixed points may be generated by running the ‘‘stable\_points.py’’ script in \url{https://github.com/boettiger-lab/approx-model-or-approx-soln}.
}
\begin{align*}
  (X,Y,Z) \approx 
  &(0.839, 0.054, 0.377), \quad 
    (0.048, 0.774, 0.480), \quad
    (0.359, 0.491, 0.509), \\
    &(0.899, 0.000, 0.303), \quad
    (0.000, 0.818, 0.457), \quad
    (0.666, 0.666, 0.000), \\
    &(1.000,0.000,0.000),\quad (0.000,1.000,0.000),\quad (0.000,0.000,0.000).
\end{align*}
The classification of fixed points for multi-dimensional systems is generally more complicated than in one dimension.
A fixed point might be stable against perturbations in one direction, but unstable against perturbations along another.
While a detailed classification of these fixed points is beyond the scope of this work, we point out that the point $(0.839, 0.054, 0.377)$ is stable against perturbations in all directions.
We will thus use it as a starting point for our simulations.

@@

We have, up to now, presented eqs. \eqref{eq:may} and \eqref{eq:3d model} as *deterministic* dynamical models for the sake of clarity.
In practice, we will consider a stochastic contribution to these dynamics.
This way, instead of eq. \eqref{eq:3d model}, we consider the problem of harvesting the following stochastic model:
<!--
\begin{align}
  \label{eq:may stoch}
  \Delta X_t
  = L(X_t;\ r, K) - F(X_t, H; \beta,c) - h_t + \eta_t,
\end{align}
where $\{\eta_t\}_t$ are independent Gaussian variables with zero mean.
The variance $\sigma^2$ of $\eta_t$ is an additional parameter of the model.

Similarly, we use the following dynamical model in place of eq. \eqref{eq:3d model}-->
\begin{align}
  \label{eq:3d model stoch}
  \begin{split}
    \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t + \eta_{X,t},\\
    \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t + \eta_{Y,t},\\
    \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t + \eta_{Z,t}.
\end{split}
\end{align}
Here, $\{\eta_{i,t}\}$ is a set of independent Gaussian variables centered at zero with variances, respectively, $\sigma_X^2$, $\sigma_Y^2$ and $\sigma_Z^2$.

@@

For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state at time $t$ being a vector of all the fish populations, $S_t=(X_t,Y_t,Z_t)$.
At each time step, the agent observes the state $S$ and harvests $X$,
\begin{align*}
  (X_t, Y_t, Z_t) \mapsto ((1-M_X)\cdot X_t, Y_t, Z_t),
\end{align*}
where the fish mortality-from-harvest $M_X=M_X(S_t)$ is the agent's action at time $t$. 
This secures a reward of $M_X(S_t)X$.
Afterwards, the environment undergoes a time step under its natural dynamics given by \eqref{eq:3d model stoch}.
